Using IRS Data to Predict Electric Vehicle Count

### 2024-08-06 code/analysis-code/statistical_analysis.qmd
add first site iteration

# Selecting columns L to CY (indexes 11 to 102)
features_to_use = data.columns[3:-1]
print(features_to_use)

@@

scalar = StandardScaler()
scale = scalar.fit_transform(X_grouped)
print(scale)

>>>>

# Selecting columns L to CY (indexes 11 to 102)
features_to_use = data.columns[3:-1]

@@

# Scale data
scalar = StandardScaler()
scale = scalar.fit_transform(X_grouped)

### 2024-08-06 code/eda-code/eda-ev-dataset.qmd
add first site iteration

```{r}
getwd()
```

@@

```{r}
# Removing null values and outlier from MSRP:
ev_data_filtered <- na.omit(ev_data) # FIltered data with NA values removed
max_msrp_outlier <- ev_data_filtered %>%
  group_by(Electric.Vehicle.Type) %>%
  summarise(max = max(Base.MSRP)) 
max_msrp_outlier
filter_msrp_outlier <- ev_data_filtered %>%
  filter(Base.MSRP != 845000) %>%
  filter(Base.MSRP != 0)
filter_msrp_outlier
```

@@

``` {r}
# Create a scatterplot - price vs. electric vehicle range correlation

p_scatterplot<-ggplot(filter_msrp_outlier, aes(x = Electric.Range, y = Base.MSRP)) +
  geom_point() +
  labs(title = "Scatterplot - Price vs. Range Correlation",
       x = "Vehicle Range",
       y = "MSRP")
plot(p_scatterplot)
```

```{r}
#For Electric.Range:
Q1 <- quantile(ev_data_filtered$Electric.Range , 0.25, na.rm = TRUE) # First Quartile
Q3 <- quantile(ev_data_filtered$Electric.Range , 0.75, na.rm = TRUE) # Third Quartile
IQR <- Q3 - Q1 # Compute Interquartile Range
lower_bound <- Q1 - 1.5 * IQR # Compute Lower Bound based on IQR
upper_bound <- Q3 + 1.5 * IQR # Compute Upper Bound based on IQR
upper_bound
```

```{r}
# Filter out outliers:
ev_data_Filtered_2 <- ev_data_filtered[ev_data_filtered$Electric.Range > lower_bound & ev_data_filtered$Electric.Range < upper_bound, ]
ev_data_Filtered_2
```

>>>>

# Introduction
This document is an exploratory data analysis of electric vehicle data. The data was obtained from the IRS and contains information on electric vehicles, their owners, and the tax benefits they receive. The data was cleaned and processed in R and saved as a CSV file. This document will load the data, explore it, and fit a linear regression model to predict the number of electric vehicles in a given zip code.

@@

```{r}
# Removing null values and outlier from MSRP:
ev_data_filtered <- na.omit(ev_data) # FIltered data with NA values removed
max_msrp_outlier <- ev_data_filtered %>%
  group_by(Electric.Vehicle.Type) %>%
  summarise(max = max(Base.MSRP)) 
max_msrp_outlier
filter_msrp_outlier <- ev_data_filtered %>%
  filter(Base.MSRP != 845000) %>%
  filter(Base.MSRP != 0)
```

@@

### 2024-08-06 code/eda-code/eda-ev-tax.qmd
add first site iteration

---
title: "EV Exploratory Data Analysis"
date: today
output: html_document
execute: 
  error: False
  warning: False
---
**Electric Vehicle Data Analysis**

# Introduction:

This document is an exploratory data analysis of electric vehicle data. The data was obtained from the IRS and contains information on electric vehicles, their owners, and the tax benefits they receive. The data was cleaned and processed in R and saved as a CSV file. This document will load the data, explore it, and fit a linear regression model to predict the number of electric vehicles in a given zip code.

@@

merged_data.to_csv('test.csv', index=False)
```

```{python}
import matplotlib.pyplot as plt

# Scatter plots to understand relationship between predictors and independent variable
for i, feature in enumerate(features_to_use):
    plt.figure(figsize=(10,5))
    plt.scatter(X_grouped[feature], y_grouped, c="lightblue", edgecolors="blue")
    plt.title(f'{feature} vs Vehicle Count')
    plt.xlabel(feature)
    plt.ylabel('Vehicle Count')
    # plt.grid(True)
    plt.savefig(f'../../results/figures/scatterplots/{feature}_scatterplot.png')

@@

# Plot the boxplots for the log-transformed features
plot_transformed_feature_boxplots(X_grouped, X_transformed)
```

>>>>

---
title: "Log-Transform Predictors"
date: today
output: html_document
execute: 
  error: False
  warning: False
---
**Electric Vehicle Data Analysis**

# Introduction
Apply log transformation to our feature

@@

merged_data.to_csv('test.csv', index=False)
```
<!-- 
```{python}
import matplotlib.pyplot as plt

# Scatter plots to understand relationship between predictors and independent variable
for i, feature in enumerate(features_to_use):
    plt.figure(figsize=(10,5))
    plt.scatter(X_grouped[feature], y_grouped, c="lightblue", edgecolors="blue")
    plt.title(f'{feature} vs. Vehicle Count Relationship')
    plt.xlabel(feature)
    plt.ylabel('Vehicle Count')
    # plt.grid(True)
    plt.savefig(f'../../results/figures/scatterplots/{feature}_scatterplot.png')

@@

# Plot the boxplots for the log-transformed features
plot_transformed_feature_boxplots(X_grouped, X_transformed)
``` -->

### 2024-08-06 index.qmd
add first site iteration

---
title: "Using IRS Data to Predict Electric Vehicle Count"
author: Sri Lakshmi Chundru, Collin Real, Joaquin Ramirez, Seth Harris
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

@@

#### Scatterplots - Predictor's vs. Vehicle Count Relationship
![SE Health Insurance Deduction Scatterplot](../../results/figures/scatterplots/Self-employed%20health%20insurance%20deduction%20amount_scatterplot.png)

@@

#### Log-Transform Histogram
![SE Health Insurance Deduction](../../results/figures/histograms/Self-employed%20health%20insurance%20deduction%20amount_histogram.png)

#### Log-Transform Boxplot
![SE Health Insurance Deduction Boxplot](../../results/figures/boxplots/Self-employed%20health%20insurance%20deduction%20amount_boxplot.png)

@@

### Model Performance Comparison
#### XGBoost
  - RMSE: 69.21 <br/>
  - R-squared: 0.892 <br/>
  - XGBoost demonstrated strong predictive capabilities, achieving a relatively low RMSE of 71.27. The model's R-squared value of 0.91 indicates that it explains 89.2% of the variance in the vehicle count data, suggesting good model fit and predictive accuracy. <br/>
![XGBoost](../../results/output/model_results/XGBoost_Regressor.png)

#### Decision Tree
  - RMSE: 79.61<br/>
  - R-squared: 0.857<br/>
  -  The Decision Tree model had the highest RMSE among the evaluated models at 79.61, reflecting less precise predictions. Its R-squared value of 0.857 indicates that it explains 85.7% of the variance, making it the least accurate model in this comparison.
![Decision Tree](../../results/output/model_results/Decision_Tree_Regressor.png)

#### HistGradientBoosting
  -  RMSE: 76.96<br/>
  -  R-squared: 0.867<br/>
  -  The HistGradientBoosting model offered a moderate performance with an RMSE of 76.96 and an R-squared value of 0.867. While better than the Decision Tree model, it still lagged behind XGBoost and Random Forest in terms of prediction accuracy and variance explanation.
![HistGradientBoosting](../../results/output/model_results/HistGradientBoosting_Regressor.png)

#### Random Forest
  -  RMSE: 71.30<br/>
  -  R-squared: 0.885<br/>
  -  The Random Forest model outperformed all other models, achieving the lowest RMSE of 71.30 and the highest R-squared value of 0.885. These metrics indicate that the Random Forest model not only provided the most accurate predictions but also explained 88.5% of the variance in vehicle counts, making it the most reliable model in this analysis.
![Random Forest](../../results/output/model_results/Random_Forest_Regressor.png)

@@

## Variable Importance - XGBoost
!['XGBoost'](../../results/output/variable_importance/XGBoost.png)

>>>>

---
title: "Using IRS Data to Predict Electric Vehicle Count"
author: Sri Lakshmi Chundru, Collin Real, Joaquin Ramirez, Seth Harris
bibliography: assets/dataanalysis-references.bib
csl: assets/apa.csl
---

@@

#### Scatterplots - Predictor's vs. Vehicle Count Relationship
![SE Health Insurance Deduction Scatterplot](results/figures/scatterplots/Self-employed%20health%20insurance%20deduction%20amount_scatterplot.png)

@@

#### Log-Transform Histogram
![SE Health Insurance Deduction](results/figures/histograms/Self-employed%20health%20insurance%20deduction%20amount_histogram.png)

#### Log-Transform Boxplot
![SE Health Insurance Deduction Boxplot](results/figures/boxplots/Self-employed%20health%20insurance%20deduction%20amount_boxplot.png)

@@

### Model Performance Comparison
#### XGBoost
  - RMSE: 69.21 <br/>
  - R-squared: 0.892 <br/>
  - XGBoost demonstrated strong predictive capabilities, achieving a relatively low RMSE of 71.27. The model's R-squared value of 0.91 indicates that it explains 89.2% of the variance in the vehicle count data, suggesting good model fit and predictive accuracy. <br/>
![XGBoost](results/output/model_results/XGBoost_Regressor.png)

#### Decision Tree
  - RMSE: 79.61<br/>
  - R-squared: 0.857<br/>
  -  The Decision Tree model had the highest RMSE among the evaluated models at 79.61, reflecting less precise predictions. Its R-squared value of 0.857 indicates that it explains 85.7% of the variance, making it the least accurate model in this comparison.
![Decision Tree](results/output/model_results/Decision_Tree_Regressor.png)

#### HistGradientBoosting
  -  RMSE: 76.96<br/>
  -  R-squared: 0.867<br/>
  -  The HistGradientBoosting model offered a moderate performance with an RMSE of 76.96 and an R-squared value of 0.867. While better than the Decision Tree model, it still lagged behind XGBoost and Random Forest in terms of prediction accuracy and variance explanation.
![HistGradientBoosting](results/output/model_results/HistGradientBoosting_Regressor.png)

#### Random Forest
  -  RMSE: 71.30<br/>
  -  R-squared: 0.885<br/>
  -  The Random Forest model outperformed all other models, achieving the lowest RMSE of 71.30 and the highest R-squared value of 0.885. These metrics indicate that the Random Forest model not only provided the most accurate predictions but also explained 88.5% of the variance in vehicle counts, making it the most reliable model in this analysis.
![Random Forest](results/output/model_results/Random_Forest_Regressor.png)

@@

## Variable Importance - XGBoost
!['XGBoost'](results/output/variable_importance/XGBoost.png)
